Factoring with $n+2$ clean qubits and $n-1$ dirty qubits

### 2017-05-25 doc/paper.tex
Ancilla plural -> ancillae

\begin{abstract}
\em
We present reversible classical circuits for performing various arithmetic operations aided by dirty ancilla in an unknown state that must be preserved.
We improve the number of clean qubits needed to factor an $n$-bit number with Shor's algorithm \cite{Shor1999} from $2n+2$ \cite{takahashi2006, haner2016} to $n+2$, and the total number of qubits needed from $2n+2$ to $2n+1$, without increasing the asymptotic size or depth of the circuit.
\end{abstract}

@@

When constructing quantum circuits, or classical reversible circuits, an important resource is the number of available ancilla.
An ancilla is an extra bit or qubit that is available for use by a circuit as temporary workspace.
Ancilla may be initialized to a known state (``clean bit"), or be given to the circuit in an unknown state (``dirty bit") that must be restored before the circuit finishes.
Clean bits are more valuable, allowing for simpler and more compact circuit constructions, but dirty bits are more plentiful, since any temporarily unused bit is a borrowable dirty bit.

One part of a circuit can borrow dirty bits from another part of the same circuit, so circuit constructions that require only dirty bits are easier to apply under tight space constraints, or on circuit topologies where other ancilla are too far away to be acquired quickly.
When attempting to reduce the number of bits or qubits required by a circuit, converting clean ancilla to dirty ancilla is a useful intermediate goal.
In this paper, we reduce the number of qubits required to find the period of exponentiation, modulo an $n$-bit modulus, from $2n+2$ clean qubits \cite{takahashi2006, haner2016} to $n+2$ clean qubits and $n-1$ dirty qubits.
We do so without increasing the asymptotic costs, though the constant factors of our constructions are worse.

Our paper is structured as follows.
Section \ref{sec:introduction} introduces and describes the conventions our circuit constructions will follow.
In section \ref{sec:construct} we describe the circuit constructions we use to reduce period finding into constant-sized gates, while tracking the number of required dirty ancilla.
Then, in section \ref{sec:costs}, we discuss the novelty and comparative costs of the presented circuit constructions.
Finally, section \ref{sec:conclusion} concludes with a discussion of possible future improvements.

All constructions in this paper use a 2s-complement representation of integers, and all diagrams order qubits from least significant bit (LSB) at the top to most significant bit (MSB) at the bottom.
Nearly all presented circuit constructions are classical, and the required dirty qubit counts we note and show in diagrams are in terms of those classical constructions.
(The quantum trick shown in figure \ref{fig:bootstrap-ancilla} can cut dirty ancilla from several of our constructions, but only in places that don't affect the overall total for period-finding.)

@@

\begin{figure}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding.png}
  }
  \caption{\em
	High-level period finding circuit \cite{Shor1999}.
	$R$ is the modulus being queried, $B$ is a randomly chosen base, $n$ is the number of bits needed to store $R$, and $p \in \Theta(n)$ controls the precision of the phase estimation step.
    The triangles indicate how many ancilla are needed ``behind the scenes", by our constructions, to perform an operation.
	Recovering the period requires classical post-processing of the sampled output.
  }
  \label{fig:period-finding}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply.png}
  \caption{\em
    Controlled modular bimultiplication, with a constant multiplier $K$, using three modular scaled-accumulates and a swap.
    $K$ must have a multiplicative inverse modulo $R$.
    Uses no ancilla, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-multiply}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply-accumulate.png}
  \caption{\em
    Reducing a controlled modular scale-accumulate, with a constant multiplier $K$, into the modular equivalent of shift-and-add.
    Requires the modulus $R$ to be odd.
    Uses no ancilla, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-scale-accumulate}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-double.png}
  \caption{\em
    Controlled modular doubling in $O(c + n \lg n)$ gates, $O(c + n)$ depth, with 2 dirty ancilla.
    $R$ must be odd.
    Uses 1 dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:modular-double}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-addition.png}
  \caption{\em
    Controlled modular addition construction based on pivot-flips.
    Uses $2-c$ dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-add}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-offset.png}
  \caption{\em
    Controlled modular offset construction based on pivot-flips.
    Uses $2$ dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-offset}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/negate-mod.png}
  \caption{\em
    Controlled modular negation.
    Uses $2$ dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:negate-mod}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-pivot-flip.png}
  \caption{\em
    Controlled pivot-flip circuit with an enregistered pivot.
    Uses $2$ dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-pivot-flip}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-const-pivot-flip.png}
  \caption{\em
    Controlled pivot-flip circuit with a constant pivot.
    Uses $2$ dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-const-pivot-flip}
\end{figure}

@@

Comparison operations toggle a target bit based on the relationship between two input registers.
We implement comparisons as in \cite{takahashi2005}, using an addition followed by a slightly smaller subtraction that clears all changes except the overflow signal into the target bit.
For comparisons against another register this uses $1$ dirty ancilla (if controlled, otherwise no ancilla), $O(c + n)$ gates, and $O(c + n)$ depth.
For comparisons against a constant, the number of gates increases to $O(c + n \lg n)$ and the dirty ancilla is always required.

@@

\cite{takahashi2005} provides a reversible adder circuit that uses $O(n)$ gates, $O(n)$ depth, and no ancilla.
We show an equivalent circuit in figure \ref{fig:inlineadder}.
Because the construction uses the input register as workspace, it doesn't work when the input is a compile-time constant or when the target register is larger than the input register.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder.png}
  \caption{\em Adder for input and target registers of the same size.
  Requires no ancilla, and uses $O(N)$ gates and depth.
  Based on \cite{van2004, takahashi2005}.}
  \label{fig:inlineadder}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder-into-large.png}
  \caption{\em
      Adder with target larger than source, using no ancilla.
      Uses $O(n)$ gates and depth.
      The increment and decrement gates all have at least one free wire to borrow as a dirty ancilla.
  }
  \label{fig:inline-adder-into-large}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-addition.png}
  \caption{\em
  	Reducing controlled addition to uncontrolled addition.
  	Uses $O(c + n)$ gates, $O(c + n)$ depth, and a dirty ancilla.
  	When applied to offset gates, uses $O(c + n \lg n)$ gates, $O(n + c)$ depth, and two dirty ancilla.
  }
  \label{fig:controlled-addition}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/increment-many-dirty.png}
  \caption{\em Subtracting $x$ and $-x-1$ from a register increments it. Requires $O(n)$ depth, size, and $n$ dirty ancilla.}
  \label{fig:increment-many-dirty}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-bit-rotate.png}
  \caption{\em
    A controlled bit rotation / bit swap is three controlled bit-reverses.
    Uses no ancilla, $O(n+c)$ gates, and $O(n+c)$ depth.
  }
  \label{fig:bit-rotate}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-reverse.png}
  \caption{\em
    An inline controlled bit order reversal.
    Each XOR operation is a series of independent CNOTs (note that the inputs have opposite endian-ness to the outputs).
    For reversals on an even number of bits, swap the center two bits beforehand then use one of them as the center workspace bit from this construction.
    Uses no ancilla, $O(c + n)$ gates, and $O(c + n)$ depth.
  }
  \label{fig:bit-reverse}
\end{figure}

@@

Several of our constructions have used CNOTs with many controls and many targets.
To avoid paying the overhead of $c$ controls for every target, we use one big CNOT to toggle-control all the others.
We then use the available dirty ancilla to reduce the single remaining CNOT with many controls into Toffoli gates.
See figures \ref{fig:multi-not} and \ref{fig:cnot-reduction}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/multi-cnot-to-single-cnots.png}
  \caption{\em
    Separating controlled not gates with many controls and many targets into single-target single-control not-gates and one single-target many-control not-gate.
    Uses no ancilla, $O(c + n)$ gates, and $O(c + \lg n)$ depth.
  }
  \label{fig:multi-not}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/cnot-reduction.png}
  \caption{\em
    When $c-2$ dirty ancilla are available, a controlled-not with $c$ controls can be reduced into $4c - 8$ Toffoli gates.
  }
  \label{fig:cnot-reduction}
\end{figure}

@@

Previous modular addition constructions worked by temporarily storing an is-wraparound-needed comparison in a clean ancilla \cite{takahashi2006, haner2016}.
Pivot flips also require ancilla (two, actually), but the ancilla can be dirty and, in the context of period-finding, there's always qubits available to borrow whenever a pivot-flip is needed.
This improvement saves a qubit in the overall circuit, reducing the total number of qubits required for period finding (on modular exponentiation against an $n$-bit modulus) from $2n+2$ to $2n+1$.

@@

Previous published incrementers (not counting an unpublished version of our construction \cite{gidney2015} cited by \cite{haner2016}) required either $O(n^2)$ gates or $\omega(1)$ ancilla \cite{draper2000, barenco1995}.
Our classical incrementer constructions use $O(n)$ gates and a single dirty ancilla.
Note that, for classical reversible computation, 1 dirty ancilla is the minimum possible for an incrementer.
The parity of the state permutation performed by an increment operation is odd, but the parity of the permutation performed by any classical gate that doesn't cover the entire circuit is even, and odd-parity permutations can't be implemented by composing even-parity permutations.
When quantum operations are available, the final ancilla can be cut without increasing the asymptotic cost (see figure \ref{fig:bootstrap-ancilla}).

@@

Although the asymptotic costs of our constructions match previous work, the constant factors are significantly worse.
For example, our equivalent of a controlled 32-qubit multiplication uses roughly 1.3 million Toffoli gates (the specific number depends on the factor, the modulus, whether or not a number of optimizations are used, and whether or not those optimizations apply to a given case).
This cost is over an order of magnitude worse than previous work \cite{haner2016}.
The reason for the increased cost is that using dirty ancilla results in more uncomputation work, in repeating some operations twice and conditionally to do them once unconditionally, and in less efficient constructions in general.
These inefficiencies stack multiplicatively when one construction uses another as a subroutine.

@@

Of the $2n+1$ qubits our period-finding construction requires: $1$ is used for the phase estimation qubit, $n$ are used to store the work register, and $n$ are ancilla used to implement modular multiplication in terms of scaled modular addition.
All the other constructions, for simpler arithmetic, can be done with two or fewer dirty ancilla.
Furthermore, because scaled modular addition is reduced into operations that have a $\Theta(n)$ surplus of unused qubits, in context there's always more than enough dirty ancilla available to implement the simpler arithmetic inline.
There's so much slack in the simpler arithmetic constructions that improving the number of qubits used by period-finding effectively means improving the very first step, reducing multiplication into some other operation, and nothing else.

For example, consider a situation where we could cut multiplication operations into two pieces.
If we knew the factorization $p \cdot q$ of the modulus $R$, with $p$ and $q$ each having size $\approx n/2$, then the Chinese remainder theorem guarantees we could uniquely represent $x \mod R$ as the pair $(x \mod p, x \mod q)$.
Multiplications of the half-sized values could use and re-use the same half-sized clean ancilla register, so there would be three half-sized registers (one for $x \mod p$, one for $x \mod q$, and one ancilla register) for a total qubit count of $\approx 1.5n$.
Despite the large reduction in the number of ancilla available to the underlying arithmetic operations, each operation would still have more than enough ancilla and contribute no additional ancilla to the overall circuit.
(Note: because the initial value $x_0 = 1$ is trivial, and the final value is measured or discarded, it's not necessary to implement circuits that translate between the Chinese-remainder representation and the usual 2s-complement representation.)
Of course, since we use period-finding to compute the factorization of $R$, using the factorization of $R$ to optimize period-finding would be paradoxical.

>>>>

\begin{abstract}
\em
We present reversible classical circuits for performing various arithmetic operations aided by dirty ancillae in an unknown state that must be preserved.
We improve the number of clean qubits needed to factor an $n$-bit number with Shor's algorithm \cite{Shor1999} from $2n+2$ \cite{takahashi2006, haner2016} to $n+2$, and the total number of qubits needed from $2n+2$ to $2n+1$, without increasing the asymptotic size or depth of the circuit.
\end{abstract}

@@

When constructing quantum circuits, or classical reversible circuits, an important resource is the number of available ancillae.
An ancilla is an extra bit or qubit that is available for use by a circuit as temporary workspace.
Ancillae may be initialized to a known state (``clean bit"), or be given to the circuit in an unknown state (``dirty bit") that must be restored before the circuit finishes.
Clean bits are more valuable, allowing for simpler and more compact circuit constructions, but dirty bits are more plentiful, since any temporarily unused bit is a borrowable dirty bit.

One part of a circuit can borrow dirty bits from another part of the same circuit, so circuit constructions that require only dirty bits are easier to apply under tight space constraints, or on circuit topologies where other ancillae are too far away to be acquired quickly.
When attempting to reduce the number of bits or qubits required by a circuit, converting clean ancillae to dirty ancillae is a useful intermediate goal.
In this paper, we reduce the number of qubits required to find the period of exponentiation, modulo an $n$-bit modulus, from $2n+2$ clean qubits \cite{takahashi2006, haner2016} to $n+2$ clean qubits and $n-1$ dirty qubits.
We do so without increasing the asymptotic costs, though the constant factors of our constructions are worse.

Our paper is structured as follows.
Section \ref{sec:introduction} introduces and describes the conventions our circuit constructions will follow.
In section \ref{sec:construct} we describe the circuit constructions we use to reduce period finding into constant-sized gates, while tracking the number of required dirty ancillae.
Then, in section \ref{sec:costs}, we discuss the novelty and comparative costs of the presented circuit constructions.
Finally, section \ref{sec:conclusion} concludes with a discussion of possible future improvements.

All constructions in this paper use a 2s-complement representation of integers, and all diagrams order qubits from least significant bit (LSB) at the top to most significant bit (MSB) at the bottom.
Nearly all presented circuit constructions are classical, and the required dirty qubit counts we note and show in diagrams are in terms of those classical constructions.
(The quantum trick shown in figure \ref{fig:bootstrap-ancilla} can cut dirty ancilla(e) from several of our constructions, but only in places that don't affect the overall total for period-finding.)

@@

\begin{figure}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding.png}
  }
  \caption{\em
	High-level period finding circuit \cite{Shor1999}.
	$R$ is the modulus being queried, $B$ is a randomly chosen base, $n$ is the number of bits needed to store $R$, and $p \in \Theta(n)$ controls the precision of the phase estimation step.
    The triangles indicate how many ancillae are needed ``behind the scenes", by our constructions, to perform an operation.
	Recovering the period requires classical post-processing of the sampled output.
  }
  \label{fig:period-finding}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply.png}
  \caption{\em
    Controlled modular bimultiplication, with a constant multiplier $K$, using three modular scaled-accumulates and a swap.
    $K$ must have a multiplicative inverse modulo $R$.
    Uses no ancillae, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-multiply}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply-accumulate.png}
  \caption{\em
    Reducing a controlled modular scale-accumulate, with a constant multiplier $K$, into the modular equivalent of shift-and-add.
    Requires the modulus $R$ to be odd.
    Uses no ancillae, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-scale-accumulate}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-double.png}
  \caption{\em
    Controlled modular doubling in $O(c + n \lg n)$ gates, $O(c + n)$ depth, with 2 dirty ancillae.
    $R$ must be odd.
    Uses 1 dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:modular-double}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-addition.png}
  \caption{\em
    Controlled modular addition construction based on pivot-flips.
    Uses $2-c$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-add}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-offset.png}
  \caption{\em
    Controlled modular offset construction based on pivot-flips.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-offset}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/negate-mod.png}
  \caption{\em
    Controlled modular negation.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:negate-mod}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-pivot-flip.png}
  \caption{\em
    Controlled pivot-flip circuit with an enregistered pivot.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-pivot-flip}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-const-pivot-flip.png}
  \caption{\em
    Controlled pivot-flip circuit with a constant pivot.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-const-pivot-flip}
\end{figure}

@@

Comparison operations toggle a target bit based on the relationship between two input registers.
We implement comparisons as in \cite{takahashi2005}, using an addition followed by a slightly smaller subtraction that clears all changes except the overflow signal into the target bit.
For comparisons against another register this uses $1$ dirty ancilla (if controlled, otherwise no ancillae), $O(c + n)$ gates, and $O(c + n)$ depth.
For comparisons against a constant, the number of gates increases to $O(c + n \lg n)$ and the dirty ancilla is always required.

@@

\cite{takahashi2005} provides a reversible adder circuit that uses $O(n)$ gates, $O(n)$ depth, and no ancillae.
We show an equivalent circuit in figure \ref{fig:inlineadder}.
Because the construction uses the input register as workspace, it doesn't work when the input is a compile-time constant or when the target register is larger than the input register.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder.png}
  \caption{\em Adder for input and target registers of the same size.
  Requires no ancillae, and uses $O(N)$ gates and depth.
  Based on \cite{van2004, takahashi2005}.}
  \label{fig:inlineadder}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder-into-large.png}
  \caption{\em
      Adder with target larger than source, using no ancillae.
      Uses $O(n)$ gates and depth.
      The increment and decrement gates all have at least one free wire to borrow as a dirty ancilla.
  }
  \label{fig:inline-adder-into-large}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-addition.png}
  \caption{\em
  	Reducing controlled addition to uncontrolled addition.
  	Uses $O(c + n)$ gates, $O(c + n)$ depth, and a dirty ancilla.
  	When applied to offset gates, uses $O(c + n \lg n)$ gates, $O(n + c)$ depth, and two dirty ancillae.
  }
  \label{fig:controlled-addition}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/increment-many-dirty.png}
  \caption{\em Subtracting $x$ and $-x-1$ from a register increments it. Requires $O(n)$ depth, size, and $n$ dirty ancillae.}
  \label{fig:increment-many-dirty}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-bit-rotate.png}
  \caption{\em
    A controlled bit rotation / bit swap is three controlled bit-reverses.
    Uses no ancillae, $O(n+c)$ gates, and $O(n+c)$ depth.
  }
  \label{fig:bit-rotate}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-reverse.png}
  \caption{\em
    An inline controlled bit order reversal.
    Each XOR operation is a series of independent CNOTs (note that the inputs have opposite endian-ness to the outputs).
    For reversals on an even number of bits, swap the center two bits beforehand then use one of them as the center workspace bit from this construction.
    Uses no ancillae, $O(c + n)$ gates, and $O(c + n)$ depth.
  }
  \label{fig:bit-reverse}
\end{figure}

@@

Several of our constructions have used CNOTs with many controls and many targets.
To avoid paying the overhead of $c$ controls for every target, we use one big CNOT to toggle-control all the others.
We then use the available dirty ancillae to reduce the single remaining CNOT with many controls into Toffoli gates.
See figures \ref{fig:multi-not} and \ref{fig:cnot-reduction}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/multi-cnot-to-single-cnots.png}
  \caption{\em
    Separating controlled not gates with many controls and many targets into single-target single-control not-gates and one single-target many-control not-gate.
    Uses no ancillae, $O(c + n)$ gates, and $O(c + \lg n)$ depth.
  }
  \label{fig:multi-not}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/cnot-reduction.png}
  \caption{\em
    When $c-2$ dirty ancillae are available, a controlled-not with $c$ controls can be reduced into $4c - 8$ Toffoli gates.
  }
  \label{fig:cnot-reduction}
\end{figure}

@@

Previous modular addition constructions worked by temporarily storing an is-wraparound-needed comparison in a clean ancilla \cite{takahashi2006, haner2016}.
Pivot flips also require ancillae (two, actually), but the ancillae can be dirty and, in the context of period-finding, there's always qubits available to borrow whenever a pivot-flip is needed.
This improvement saves a qubit in the overall circuit, reducing the total number of qubits required for period finding (on modular exponentiation against an $n$-bit modulus) from $2n+2$ to $2n+1$.

@@

Previous published incrementers (not counting an unpublished version of our construction \cite{gidney2015} cited by \cite{haner2016}) required either $O(n^2)$ gates or $\omega(1)$ ancillae \cite{draper2000, barenco1995}.
Our classical incrementer constructions use $O(n)$ gates and a single dirty ancilla.
Note that, for classical reversible computation, 1 dirty ancilla is the minimum possible for an incrementer.
The parity of the state permutation performed by an increment operation is odd, but the parity of the permutation performed by any classical gate that doesn't cover the entire circuit is even, and odd-parity permutations can't be implemented by composing even-parity permutations.
When quantum operations are available, the final ancilla can be cut without increasing the asymptotic cost (see figure \ref{fig:bootstrap-ancilla}).

@@

Although the asymptotic costs of our constructions match previous work, the constant factors are significantly worse.
For example, our equivalent of a controlled 32-qubit multiplication uses roughly 1.3 million Toffoli gates (the specific number depends on the factor, the modulus, whether or not a number of optimizations are used, and whether or not those optimizations apply to a given case).
This cost is over an order of magnitude worse than previous work \cite{haner2016}.
The reason for the increased cost is that using dirty ancillae results in more uncomputation work, in repeating some operations twice and conditionally to do them once unconditionally, and in less efficient constructions in general.
These inefficiencies stack multiplicatively when one construction uses another as a subroutine.

@@

Of the $2n+1$ qubits our period-finding construction requires: $1$ is used for the phase estimation qubit, $n$ are used to store the work register, and $n$ are ancillae used to implement modular multiplication in terms of scaled modular addition.
All the other constructions, for simpler arithmetic, can be done with two or fewer dirty ancillae.
Furthermore, because scaled modular addition is reduced into operations that have a $\Theta(n)$ surplus of unused qubits, in context there's always more than enough dirty ancillae available to implement the simpler arithmetic inline.
There's so much slack in the simpler arithmetic constructions that improving the number of qubits used by period-finding effectively means improving the very first step, reducing multiplication into some other operation, and nothing else.

For example, consider a situation where we could cut multiplication operations into two pieces.
If we knew the factorization $p \cdot q$ of the modulus $R$, with $p$ and $q$ each having size $\approx n/2$, then the Chinese remainder theorem guarantees we could uniquely represent $x \mod R$ as the pair $(x \mod p, x \mod q)$.
Multiplications of the half-sized values could use and re-use the same half-sized clean ancilla register, so there would be three half-sized registers (one for $x \mod p$, one for $x \mod q$, and one ancilla register) for a total qubit count of $\approx 1.5n$.
Despite the large reduction in the number of ancillae available to the underlying arithmetic operations, each operation would still have more than enough ancillae and contribute no additional ancillae to the overall circuit.
(Note: because the initial value $x_0 = 1$ is trivial, and the final value is measured or discarded, it's not necessary to implement circuits that translate between the Chinese-remainder representation and the usual 2s-complement representation.)
Of course, since we use period-finding to compute the factorization of $R$, using the factorization of $R$ to optimize period-finding would be paradoxical.
